\documentclass{article}
% Language setting
% Replace `english' with e.g. `spanish' to change the document language
\usepackage[english]{babel}

% Set page size and margins
% Replace `letterpaper' with `a4paper' for UK/EU standard size
\usepackage[a4paper,margin=0.5in]{geometry}

% Useful packages
\usepackage{amsmath,amsfonts, caption}
\usepackage{graphicx,float,booktabs}
\usepackage{authblk}
\usepackage{float}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{natbib}
\usepackage{indentfirst}
\usepackage{lipsum}

\title{Estimating Somatic Variant Quality from Clonal Sequencing using Hierarchical Bayesian Models}
\author[1,*]{Arjun Biddanda}
\affil[1]{Johns Hopkins University, Department of Biology}
\affil[*]{Contact: \href{mailto:abiddan1@jhu.edu}{abiddan1@jhu.edu}}

\date{\today}

\begin{document}

\maketitle


\begin{abstract}
The detection of \textit{somatic} variants from ``clonal'' sampling of tissues is a major challenge in leveraging sequencing data to understand both tissue development and cancer progression. Detecting truly somatic variants requires disentangling sequencing error and true germline polymorphism, often requiring high sequencing depth per-clone ($> 10x$ average depth). Additionally, there may be high variation in sequencing depths \textit{per clone}, which affects the power to detect a somatic variant, particularly at low variant allele frequencies (VAF). Furthermore, if a germline reference sample is provided but is not of high quality, there may be uncertainty even for high-VAF somatic variants. To address the challenge of somatic variant filtering and curation, we develop a hierarchical model leveraging multiple annotations from publicly available datasets to estimate the \textit{posterior probability of germline polymorphism} as a comparable statistic across multiple clonally sampled individuals for consistent filtering. We highlight how variation in depth of coverage amongst clones and sequencing errors can affect somatic variant detection power. Our method provides variant-level annotations for filtering across variety of somatic sequencing contexts.
\end{abstract}

\section*{Problem Motivation}

Our goal here is to develop a method to estimate the \textit{probability} that an observed clonal variant is actually somatic and not a germline variant. In actuality, we will be estimating the ``complement'', the probability that the variant is actually a germline variant. The core intuition used by a number of somatic variant callers is that (1) a germline variant is shared across \textit{all} somatic clones and (2) is also present in a germline sample \citep{Coorens2021-xy,Coorens2024-be,Spencer-Chapman2025-pj}. Without loss of generality we assume that ``clone'' refers to sequencing data from a single colony derived from a blood progenitor lineage or in cases of other tissues, a laser-capture microdissection.

We specifically want to model how the germline sample, alongside additional information can provide ``prior'' information on whether a variant is likely to be polymorphic in the germline (e.g., population allele frequency). The likelihood is then defined using the collected evidence for that polymorphism across \textit{all} of the individual clone sequences. The prior probability of germline polymorphism will be informed by specific variant-level annotations (e.g., variant-level masks, allele frequencies, region-specific mutation rate). If the hyperparameters dictating the prior distribution are estimated using Empirical Bayes, we can then robustly arrive at a posterior probability of germline polymorphism estimate per putative somatic variant.

\section*{Model}

We start from a model where the output consists of reference and alternative read counts from each of the $J$ clonal samples at the $k^{th}$ site ($R_k$ and $A_k$ respectively). Following previous modeling, we will consider that the $k^{th}$ variant falls in a category $z_k\in \{\texttt{somatic}, \texttt{het}\}$, where $\texttt{het}$ is a germline heterozygote. Specifically, assuming that the $J$ clones are independent, we can model the pair of read-counts ($a_{jk}, r_{jk}$), where $n_{jk} = a_{jk} + r_{jk}$, conditional on $z_k$:


\begin{equation}
P(a_{jk}, r_{jk} \mid z_k = \texttt{het}) = \texttt{Bin}(a_{jk},n_{jk},\frac{1}{2})
\label{eq:germline_ll}
\end{equation}


\begin{equation}
P(a_{jk}, r_{jk} \mid z_k = \texttt{somatic},\phi_j,\mu,\kappa )
= \phi_j \cdot \texttt{Bin}\left(a_{jk},n_{jk},\frac{1}{2}\right) + (1 - \phi_j)\cdot \texttt{BetaBin}(a_{jk},n_{jk}; \mu\kappa, (1-\mu)\kappa)
\label{eq:somatic_ll}
\end{equation}

where $\mu = 10^{-3}$ is the prior mean error rate and $\kappa > 0$ is
the concentration parameter estimated from data via the EM-algorithm. The Beta-Binomial distribution in Eq.\ref{eq:somatic_ll} comes from integration of alternative reads arise from sequencing error whose rate is drawn from $\texttt{Beta}(\mu\kappa,(1{-}\mu)\kappa)$, integrating over site-to-site variability in error.

Combining these likelihoods across all clones, we obtain the following overall likelihoods for the $k^{th}$ site:

\begin{equation}
P(\mathbf{A}_k, \mathbf{R}_k | z_k=\texttt{het}) = \prod_j^J  \texttt{Bin}\left(a_{jk},n_{jk},\frac{1}{2}\right)
\label{eq:full_germline_logll}
\end{equation}

and for a somatic variant,

\begin{equation}
P(\mathbf{A}_k, \mathbf{R}_k | z_k=\texttt{somatic}) = \prod_j^J  \phi_j \cdot \texttt{Bin}\left(a_{jk},n_{jk},\frac{1}{2}\right) + (1 - \phi_j)\cdot \texttt{BetaBin}(a_{jk},n_{jk}; \mu\kappa, (1-\mu)\kappa)
\label{eq:full_somatic_logll}
\end{equation}


\subsection*{Annotation-Informed Prior on $\phi_j$}

The carrier probability $\phi_j$ is the key parameter distinguishing true germline variants from somatic variants, and it need not be a single shared constant. Site-level and clone-level annotations can push $\phi_j$ toward 1 (strong
evidence the clone carries the mutation) or toward 0 (strong evidence it does not). We propose a logistic regression model:

\begin{equation}
  \mathrm{logit}(\phi_j) =  \mu_k + \delta_j =
  \sum_{\ell} \lambda_\ell \Theta_{k\ell} + \sum_{m} \beta_m \Phi_{jm}
  \label{eq:logistic}
\end{equation}

, where $\Theta_{k\ell}$ are \emph{site-level} annotations (shared across all clones at site $k$) and $\Phi_{jm}$ are \emph{clone-level} annotations (specific to clone $j$).  The two additive terms $\mu_k$ and $\delta_j$ correspond to the site's intrinsic propensity to be somatic and the clone's pecific evidence for carrying it, respectively.


\subsubsection*{Site-Level Annotations and Potential Effects on $\phi_j$}

These annotations shift $\phi_j$ for \emph{all} clones at a site, because
they reflect properties of the variant itself rather than of any individual
clone.

\paragraph{Population allele frequency (gnomAD AF).}
If a variant is observed at high frequency in population databases (e.g.\
gnomAD AF $> 0.001$), it is almost certainly a germline polymorphism rather
than a somatic mutation, so $\phi_j \to 0$ for all $j$.  The relevant
annotation is $\Theta_{k,\mathrm{AF}} = \log_{10}(\mathrm{AF}_k + \delta)$ for
a small offset $\delta$ to handle AF $= 0$.  A large negative coefficient
$\lambda_\mathrm{AF} \ll 0$ would correctly suppress $\phi_j$ at high-AF
sites.  Conversely, a variant absent from gnomAD provides evidence in favour of
somatic origin.

\paragraph{Regional mappability and repetitive elements.}
Low-mappability regions, segmental duplications, and simple repeats produce
alignment artefacts that elevate apparent VAF in all clones simultaneously.
Including a mappability score (e.g.\ from ENCODE) or a binary repeat-mask flag
as $\Theta_{k,\mathrm{map}}$ allows the model to globally suppress $\phi_j$
at problematic loci.

\paragraph{Local GC content and trinucleotide context.}
The per-base error rate $\epsilon_{jk}$ covaries with GC content and
sequence context.  Additionally, certain trinucleotide contexts are enriched
for specific mutational signatures (e.g.\ C$\to$T at CpG for clock-like
mutations, C$\to$A for oxidative damage).  A site embedded in a high-prior
mutational signature context would have a higher baseline $\phi_j$; one in a
known artefact context (e.g.\ OxoG at CAC/CTC) would be suppressed.

\paragraph{Germline sample genotype likelihood (GL).}
If a matched germline reference is available, the phred-scaled genotype
likelihoods $\mathrm{GL}(0/0)$, $\mathrm{GL}(0/1)$, $\mathrm{GL}(1/1)$ can
directly inform the prior.  A high $\mathrm{GL}(0/1)$ in the germline sample is
strong evidence that $z_k = \mathrm{het}$, globally suppressing $\phi_j \to
0$.  Crucially, the \emph{uncertainty} of the GL (i.e.\ low depth in the
germline sample) should be propagated: rather than using the called genotype,
the raw GL values should enter the logistic model, allowing the model to
discount uncertain germline calls naturally.

\subsection*{Clone-Level Annotations impacting $\phi_j$ Across Clones}

These annotations vary across clones at the same site, allowing the model to
assign different carrier probabilities to different clones.

\paragraph{Clone sequencing depth ($n_{jk}$).}
A clone sequenced at very low depth at site $k$ has poor power to confirm or
refute carrier status.  The posterior $\gamma_{jk}$ (Section~\ref{sec:em})
should be shrunk toward the prior $\phi_j$ in this case.  Formally, depth
enters the likelihood directly (it determines how informative $a_{jk}$ is), but
it can also be included as a covariate $\Phi_{j,\mathrm{depth}} = \log n_{jk}$
to regularise $\delta_j$ for low-depth clones.

\paragraph{Clone-level strand bias and base quality.}
If alt reads in clone $j$ are predominantly on one strand or have low base
quality scores, this is evidence of an artefact specific to that clone,
suppressing $\phi_j$ for that clone alone.  Strand bias (e.g.\ Fisher's
exact $p$-value on strand counts) and mean base quality of alt reads can be
included as clone-level covariates.

\paragraph{Clone-level copy number state.}
If clone $j$ has a copy number alteration at locus $k$ (detectable from
read-depth variation or allele-specific coverage), the expected VAF within that
clone deviates from $\tfrac{1}{2}$, making the $\texttt{Bin}(\cdot;
\tfrac{1}{2})$ component inappropriate.  A copy-number state annotation
$\Phi_{j,\mathrm{CN}}$ can either suppress $\phi_j$ to flag the clone
as unreliable for this site.

\subsubsection*{Summary of Annotation Effects}

\begin{center}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{|c|c|c|c|c|}
\toprule
Annotation & Level & Effect on $\phi_j$ & Direction \\
\midrule
gnomAD AF & Site & All clones & $\uparrow$ AF $\Rightarrow$ $\phi_j \downarrow$ \\
Germline GL(0/1) & Site & All clones & $\uparrow$ GL $\Rightarrow$ $\phi_j \downarrow$ \\
CADD / driver gene & Site & All clones & $\uparrow$ score $\Rightarrow$ $\phi_j \uparrow$ \\
Mappability / repeat mask & Site & All clones & Low map $\Rightarrow$ $\phi_j \downarrow$ \\
Mutational signature context & Site & All clones & Signature match $\Rightarrow$ $\phi_j \uparrow$ \\
Clone depth $n_{jk}$ & Clone & Clone $j$ only & Low depth $\Rightarrow$ shrink to prior \\
Strand bias (clone $j$) & Clone & Clone $j$ only & High bias $\Rightarrow$ $\phi_j \downarrow$ \\
Copy number (clone $j$) & Clone & Clone $j$ only & CNA $\Rightarrow$ $\phi_j$ unreliable \\
\bottomrule
\end{tabular}
\end{center}

\subsection*{Beta Prior on the Sequencing Error Rate}

Without external information, the per-site sequencing error rate
$\epsilon_{jk}$. We instead treat it as a latent variable and marginalise it out analytically.  We place a Beta prior on $\epsilon_{jk}$ with fixed mean $\mu$ and learnable concentration $\kappa$:

\begin{equation}
  \epsilon_{jk} \;\sim\; \mathrm{Beta}(\mu\kappa,\;(1-\mu)\kappa),
  \qquad \mu = 10^{-3} \text{ (fixed)}, \quad \kappa > 0 \text{ (estimated)}
  \label{eq:epsprior}
\end{equation}

The mean $\mu = 10^{-3}$ reflects the typical error rate of  Illumina short-read sequencing and is treated as a platform constant, not a free parameter. Estimating $\mu$ from data would risk the error component absorbing genuine low-VAF somatic signal.

The concentration $\kappa$ controls how tightly $\epsilon_{jk}$ is pinned to
$\mu$:
\begin{itemize}
  \item $\kappa \to \infty$: all sites share exactly $\epsilon = \mu$
        (no site-to-site variability).
  \item $\kappa \sim 10$: wide variance; $\epsilon_{jk}$ can range from near
        zero to $\sim 10^{-2}$, appropriate when GC content and sequence context
        drive substantial site-level error variation.
  \item $\kappa \sim 1000$: tightly concentrated; appropriate for
        well-characterised, uniform sequencing conditions.
\end{itemize}

The prior variance is $\mathrm{Var}[\epsilon_{jk}] = \mu(1-\mu)/(\kappa+1)
\approx 10^{-3}/\kappa$ for small $\mu$, making the relationship between
$\kappa$ and error-rate precision transparent.


Integrating $\epsilon_{jk}$ out of the error component yields a closed-form
Beta-Binomial:
\begin{align}
  \int_0^1 \texttt{Bin}(a_{jk},n_{jk},\epsilon),
  \texttt{Beta}(\epsilon;,\mu\kappa,(1{-}\mu)\kappa),\mathrm{d}\epsilon
  &\;=\;
  \texttt{BetaBin}(a_{jk},n_{jk};,\mu\kappa,(1{-}\mu)\kappa)
  \label{eq:bbintegral}
\end{align}

where
\begin{equation}
  \texttt{BetaBin}(a,n;\alpha,\beta) = \binom{n}{a} \frac{B(a + \alpha,\; n - a + \beta)}{B(\alpha,\;\beta)}
\end{equation}
and $B(\cdot,\cdot)$ is the Beta function, evaluable in $O(1)$ via log-Gamma
arithmetic. This is the origin of the \texttt{BetaBin} in
Equation~\eqref{eq:somatic_ll}: it appears exclusively in the \emph{error} component.

\subsection*{Posterior Probability of Germline Heterozygosity}

\subsubsection*{The prior probability of germline heterozygosity $\pi_k$}

The prior probability that site $k$ is a germline heterozygote,
$\pi_k = P(z_k = \mathrm{het})$, should depend on \emph{all} available
annotations --- both site-level and clone-level --- because both carry
information about whether the variant is likely germline before any read
evidence is examined.  We therefore write:

\begin{equation}
  \pi_k
  \;=\;
  P\!\left(z_k = \mathrm{het} \mid \boldsymbol{\Theta}_k,\,
  \overline{\boldsymbol{\Phi}}_k,\, \boldsymbol{\lambda},\, \boldsymbol{\beta}\right)
  \;=\;
  \sigma\!\left(
    \sum_\ell \lambda_\ell \Theta_{k\ell}
    \;+\;
    \sum_m \beta_m \,\overline{\Phi}_{km}
  \right)
  \label{eq:pi}
\end{equation}

where $\sigma(x) = \frac{1}{(1 + e^{-x}}$ is the logistic sigmoid and
$\overline{\Phi}_{km} = J^{-1}\sum_j \Phi_{jm}$ is the \emph{clone-average}
of the $m$th clone-level annotation at site $k$.

% \paragraph{Why clone-level annotations enter $\pi_k$ via their average.}
% $\pi_k$ is a \emph{site-level} scalar --- the prior on $z_k$ before observing
% any reads --- so it cannot depend on clone-specific quantities directly.
% However, the mean across clones $\overline{\Phi}_{km}$ is itself a site-level
% summary that carries genuine prior information.  For example:
% \begin{itemize}
%   \item If the mean strand bias across all clones at site $k$ is high, this
%         suggests a systematic artefact at that position, increasing the prior
%         probability of germline or artefact origin and thus increasing $\pi_k$.
%   \item If mean clone depth $\overline{n}_k$ is low, the site is poorly
%         characterised across the board, and the prior should be less
%         informative (shrunk toward $\tfrac{1}{2}$); this can be encoded by
%         including $\overline{n}_k$ with a small positive coefficient
%         $\beta_{\mathrm{depth}}$.
%   \item If the mean per-clone alt read fraction across clones (a site-level
%         pooled VAF estimate) is close to $\tfrac{1}{2}$, this provides
%         \emph{prior} evidence that the variant behaves like a heterozygote
%         globally, increasing $\pi_k$.
% \end{itemize}

Importantly, $\boldsymbol{\lambda}$ and $\boldsymbol{\beta}$ are
\emph{shared} between clones affecting $\pi_k$ (Equation~\ref{eq:pi}) and $\phi_j$
(Equation~\ref{eq:logistic}). This means the same annotation weights
simultaneously govern (i) the site-level germline prior and (ii) the
clone-level carrier probability.

\subsubsection*{Full Posterior}

\begin{equation}
  P\!\left(z_k = \texttt{het} \mid \mathbf{A}_k, \mathbf{R}_k\right)
  \;=\;
  \frac{
    \pi_k \cdot P\left(\texttt{A}_k, \mathbf{R}_k \mid z_k = \texttt{het}\right)
  }{
    \pi_k \cdot P\left(\texttt{A}_k, \mathbf{R}_k \mid z_k = \texttt{het}\right)
    + (1 - \pi_k)\cdot P\left(\mathbf{A}_k, \mathbf{R}_k \mid z_k = \texttt{somatic}\right)
  }
\label{eq:posterior}
\end{equation}

where $\pi_k$ is given by Equation~\eqref{eq:pi}, $P(\mathbf{A}_k, \mathbf{R}_k \mid z_k = \mathrm{somatic})$ by
Equation~\eqref{eq:full_somatic_logll}, and $P(\mathbf{A}_k, \mathbf{R}_k \mid z_k =
\mathrm{het})$ by Equation~\eqref{eq:full_germline_logll}.  All three quantities share the parameters $(\boldsymbol{\lambda}, \boldsymbol{\beta}, \kappa)$ estimated jointly by EM.


\subsection*{EM Algorithm Definition and Computational Complexity}

The free parameters are $\boldsymbol{\theta} = (\boldsymbol{\lambda},
\boldsymbol{\beta}, \kappa)$, where $\boldsymbol{\lambda}$ and
$\boldsymbol{\beta}$ are the logistic regression weights shared between the
site-level germline prior $\pi_k$ (Equation~\ref{eq:pi}) and the clone-level
carrier probability $\varphi_j$ (Equation~\ref{eq:logistic}), and $\kappa$ is
the error-rate concentration.  The mean $\mu = 10^{-3}$ is fixed throughout.

\subsubsection*{E-Step: Responsibilities}

For each site $k$, compute the posterior probability of germline heterozygosity
(used as the site-level soft label in the M-step):
\begin{equation}
  \eta_k
  \;=\;
  \frac{
    P\!\left(\mathbf{A}_k, \mathbf{R}_k \mid z_k = \mathrm{het}\right)\,\pi_k
  }{
    P\!\left(\mathbf{A}_k, \mathbf{R}_k \mid z_k = \mathrm{het}\right)\,\pi_k
    \;+\;
    P\!\left(\mathbf{A}_k, \mathbf{R}_k \mid z_k = \mathrm{somatic}\right)\,(1-\pi_k)
  }
\end{equation}

For each clone $j$ at site $k$, compute the posterior probability that clone
$j$ carries the mutation (used as the clone-level soft label in the M-step):

\begin{equation}
  \gamma_{jk}
  \;=\;
  \frac{
    \varphi_j \cdot \mathrm{Bin}\!\left(a_{jk},\,n_{jk},\,\tfrac{1}{2}\right)
  }{
    \varphi_j \cdot \mathrm{Bin}\!\left(a_{jk},\,n_{jk},\,\tfrac{1}{2}\right)
    \;+\;
    (1 - \varphi_j)\cdot \mathrm{BetaBin}\!\left(a_{jk},\,n_{jk};\,\mu\kappa,\,(1{-}\mu)\kappa\right)
  }
  \label{eq:estep}
\end{equation}

The matrix $\{\gamma_{jk}\} \in [0,1]^{J \times K}$ is the responsibility
matrix: $\gamma_{jk} \approx 1$ signals that clone $j$ likely carries the
somatic mutation at site $k$, and is a useful by-product for clone-resolved
mutation burden estimation.

\subsubsection*{M-Step: Parameter Updates}

\paragraph{Logistic weights $(\boldsymbol{\lambda}, \boldsymbol{\beta})$.}
Because $\boldsymbol{\lambda}$ and $\boldsymbol{\beta}$ appear in both $\pi_k$
and $\varphi_j$, the M-step objective combines contributions from both levels:
\begin{align}
  \hat{\boldsymbol{\lambda}},\, \hat{\boldsymbol{\beta}}
  \;=\;
  \operatorname*{arg\,max}_{\boldsymbol{\lambda},\,\boldsymbol{\beta}}
  \Bigl\{
  &\underbrace{\sum_{k}
  \left[
    \eta_k \log\pi_k(\boldsymbol{\lambda},\boldsymbol{\beta})
    + (1-\eta_k)\log(1-\pi_k(\boldsymbol{\lambda},\boldsymbol{\beta}))
  \right]}_{\text{site-level: prior on }z_k}
  \notag \\
  +\;
  &\underbrace{\sum_{j,k}
  \left[
    \gamma_{jk}\log\varphi_j(\boldsymbol{\lambda},\boldsymbol{\beta})
    + (1-\gamma_{jk})\log(1-\varphi_j(\boldsymbol{\lambda},\boldsymbol{\beta}))
  \right]}_{\text{clone-level: carrier probability}}
  \Bigr\}
  \label{eq:mstep_lambda}
\end{align}
where $\eta_k = P(z_k = \mathrm{het} \mid \mathbf{A}_k, \mathbf{R}_k)$ is the
current posterior from Equation~\eqref{eq:posterior}, acting as a soft label
for the site-level binary regression on $z_k$.  The combined objective is again
a weighted logistic regression --- now over both sites (for $\pi_k$) and
site-clone pairs (for $\varphi_j$) --- solvable by L-BFGS.

\paragraph{Concentration parameter $\kappa$.}
The M-step for $\kappa$ maximises the expected log-likelihood of the
\emph{error component}, weighted by $(1 - \gamma_{jk})$:
\begin{equation}
  \hat{\kappa}
  \;=\;
  \operatorname*{arg\,max}_{\kappa > 0}
  \;\mathcal{Q}(\kappa)
  \;=\;
  \operatorname*{arg\,max}_{\kappa > 0}
  \sum_{j,k}
  (1 - \gamma_{jk})\,
  \log \mathrm{BetaBin}\!\left(a_{jk},\,n_{jk};\,\mu\kappa,\,(1{-}\mu)\kappa\right)
  \label{eq:mstep_kappa}
\end{equation}

This objective has no closed form.  Taking the derivative with respect to
$\kappa$ and setting to zero gives the score equation:
\begin{align}
  \frac{\mathrm{d}\mathcal{Q}}{\mathrm{d}\kappa}
  &= \sum_{j,k} (1 - \gamma_{jk})
  \Bigl[
    \mu\,\psi(a_{jk} + \mu\kappa)
    + (1{-}\mu)\,\psi(n_{jk} - a_{jk} + (1{-}\mu)\kappa) \notag \\
  &\qquad\qquad\qquad\quad
    - \psi(n_{jk} + \kappa)
    - \mu\,\psi(\mu\kappa)
    - (1{-}\mu)\,\psi((1{-}\mu)\kappa)
    + \psi(\kappa)
  \Bigr]
  \;=\; 0
  \label{eq:kappa_score}
\end{align}

where $\psi(\cdot) = \frac{\mathrm{d}}{\mathrm{d}x}\log\Gamma(x)$ is the
digamma function.  Because $\mathcal{Q}(\kappa)$ is smooth and typically
unimodal on $(0, \infty)$, Equation~\eqref{eq:kappa_score} is solved
efficiently by a \textbf{1D line search} (e.g.\ Brent's method).  Each evaluation of
$\mathcal{Q}$ costs $O(JK)$ digamma calls, each $O(1)$, so the line search
adds no change to the asymptotic complexity of the M-step.

\subsection*{Practical Notes}

\begin{itemize}
  \item \textbf{Fixing $\mu$.}  The mean $\mu = 10^{-3}$ is not updated in the
        M-step.  Allowing $\mu$ to float risks absorbing low-VAF somatic signal
        into the error component, particularly at low-depth sites.
  \item \textbf{Initialisation.}  A sensible default is $\kappa_0 = 100$
        (variance $\approx 10^{-5}$, consistent with moderate site-to-site
        error variability) and $\phi_j^{(0)} = 0.5$ for all $j$.
  \item \textbf{Convergence.}  Monitor the change in the full observed-data
        log-likelihood across iterations; typically 10--30 EM iterations
        suffice.
\end{itemize}

The responsibility matrix $\{\gamma_{jk}\}$ is a useful by-product: it
identifies which specific clones likely carry each somatic mutation, enabling
clone-resolved mutation burden estimates.


\subsection*{Computational Complexity}
\label{sec:complexity}

We analyse the time complexity of each component of the model in terms of:
\begin{itemize}
  \item $J$ --- number of clones
  \item $K$ --- number of candidate variant sites
  \item $L$ --- number of site-level annotations
  \item $M$ --- number of clone-level annotations
  \item $T$ --- number of inner optimiser iterations in the M-step logistic regression
  \item $S$ --- number of function evaluations in the Brent line search for $\kappa$
\end{itemize}

Throughout, evaluating a single Binomial PMF $\mathrm{Bin}(a, n, p)$ costs
$O(1)$ via precomputed log-factorial tables.  Evaluating a Beta--Binomial PMF
$\mathrm{BetaBin}(a, n; \alpha, \beta)$ also costs $O(1)$ via log-Gamma
arithmetic (specifically, three evaluations of $\log\Gamma$).

\subsubsection*{Per-Clone Likelihood (Equation~\ref{eq:perclone})}

At a single site $k$ and clone $j$, the mixture requires one Binomial and one
Beta--Binomial evaluation plus one weighted sum, all $O(1)$:
\begin{equation*}
  \mathcal{T}_{\text{per clone}} = O(1)
\end{equation*}
Across all $J$ clones at one site:
\begin{equation*}
  \mathcal{T}_{\text{one site}} = O(J)
\end{equation*}
Across all $K$ sites:
\begin{equation}
  \boxed{\mathcal{T}_{\text{per-clone likelihood}} = O(JK)}
\end{equation}

\subsubsection*{Full Somatic Likelihood (Equation~\ref{eq:full})}

The product over $J$ clones at each site $k$ is $O(J)$; with $K$ sites this
remains $O(JK)$.  When $\varphi_j$ is obtained directly from the logistic model
(Section~\ref{sec:annot}) rather than marginalised by numerical quadrature, no
additional cost is incurred:
\begin{equation}
  \boxed{\mathcal{T}_{\text{somatic likelihood}} = O(JK)}
\end{equation}
If instead $\varphi$ is marginalised by $Q$-point Gauss--Legendre quadrature,
the cost rises to $O(QJK)$, where $Q \approx 20$--$50$ suffices in practice.

\subsubsection*{Germline Likelihood (Equation~\ref{eq:germline})}

A single Binomial evaluation per clone per site, no mixture:
\begin{equation}
  \boxed{\mathcal{T}_{\text{germline likelihood}} = O(JK)}
\end{equation}

\subsubsection*{Logistic Model for \texorpdfstring{$\varphi_j$}{phi\_j}
  (Equation~\ref{eq:logistic})}

The site-level linear combination $\mu_k = \sum_\ell \lambda_\ell \Theta_{k\ell}$
costs $O(L)$ per site, giving $O(KL)$ across all sites.  The clone-level term
$\delta_j = \sum_m \beta_m \Phi_{jm}$ costs $O(M)$ per clone, giving $O(JM)$
per site and $O(JKM)$ across all sites and clones.  Applying the logistic
sigmoid and combining the two terms:
\begin{equation}
  \boxed{\mathcal{T}_{\varphi_j} = O\!\left(K(L + JM)\right)}
\end{equation}
When $M = 0$ (site-level annotations only) this reduces to $O(KL)$; when $L =
0$ (clone-level only) it is $O(JKM)$.

\subsubsection*{EM E-Step}

Computing $\gamma_{jk}$ for all $j, k$ requires one evaluation of the per-clone
likelihood (numerator and denominator share terms and can be computed jointly):
\begin{equation}
  \boxed{\mathcal{T}_{\text{E-step}} = O(JK)}
\end{equation}

\subsubsection*{EM M-Step}

\paragraph{Logistic regression update.}
The weighted logistic regression over $JK$ observations with $L + M$ parameters
costs $O(JK(L + M))$ per gradient or Newton step, iterated for $T$ steps of
the inner optimiser (e.g.\ L-BFGS):
\begin{equation}
  \boxed{\mathcal{T}_{\text{M-step, }\boldsymbol{\lambda}/\boldsymbol{\beta}}
  = O\!\left(T \cdot JK(L + M)\right)}
\end{equation}

\paragraph{Concentration parameter $\kappa$ update.}
Each evaluation of $\mathcal{Q}(\kappa)$ in Equation~\eqref{eq:mstep_kappa}
requires $O(JK)$ digamma calls, each $O(1)$.  Brent's method converges in
$S \sim 20$--$50$ function evaluations:
\begin{equation}
  \boxed{\mathcal{T}_{\text{M-step, }\kappa} = O(S \cdot JK)}
\end{equation}
Since $S \ll T(L+M)$ in practice, this term is dominated by the logistic
regression update and does not change the overall M-step complexity.

\subsubsection*{Posterior Probability (Equation~\ref{eq:posterior})}

Given precomputed likelihoods, the posterior at each site is a ratio of two
scalars:
\begin{equation}
  \boxed{\mathcal{T}_{\text{posterior}} = O(K)}
\end{equation}

\subsubsection*{Three-State Genotype Extension (Section~\ref{sec:genotype})}

Adding the $g_{jk} = 2$ component introduces one additional Binomial evaluation
per clone per site.  The asymptotic complexity is unchanged:
\begin{equation}
  \boxed{\mathcal{T}_{\text{three-state}} = O(JK)}
\end{equation}

\subsubsection*{Full EM Algorithm (One Iteration)}

A single EM iteration comprises the E-step and M-step.  The logistic model
evaluation dominates when $L$ or $M$ is large:
\begin{equation}
  \boxed{\mathcal{T}_{\text{EM iteration}} = O\!\left(T \cdot JK(L + M)\right)}
\end{equation}
For typical values --- $J \sim 10$--$100$ clones, $K \sim 10^4$--$10^6$ sites,
$L + M \sim 10$--$20$ annotations, $T \sim 20$ L-BFGS steps --- this is
comfortably parallelisable across sites, as each site's likelihood factors
independently.

\subsection*{Summary Table}

\begin{center}
\renewcommand{\arraystretch}{1.45}
\begin{tabular}{lll}
\toprule
Component & Expression & Dominant term \\
\midrule
Per-clone likelihood & $O(JK)$ & $K$ sites $\times$ $J$ clones \\
Full somatic likelihood & $O(JK)$ & same \\
Germline likelihood & $O(JK)$ & same \\
Logistic model ($\varphi_j$) & $O(K(L + JM))$ & clone-level annots when $M > 0$ \\
EM E-step & $O(JK)$ & same as likelihood \\
EM M-step ($\boldsymbol{\lambda}, \boldsymbol{\beta}$) & $O(T \cdot JK(L+M))$ & inner optimiser \\
EM M-step ($\kappa$, Brent) & $O(S \cdot JK)$ & $S$ digamma evaluations; dominated by above \\
Posterior probability & $O(K)$ & one ratio per site \\
\midrule
Full EM (one iteration) & $O(T \cdot JK(L+M))$ & M-step logistic regression \\
\bottomrule
\end{tabular}
\end{center}


\section*{Model Validation and Performance}

For initial validation, we simulated read-level data under the model and for $K=1000$ estimated the contribution of two annotations $\lambda_1, \lambda_2$ to the posterior probability of being a germline polymorphic site. We artificially simulated $K' = 100$ sites as true somatic variants in this setting. \\




\section*{Limitations / Model Improvement}














\section*{Model Comparisons}

\subsection*{Beta-Binomial Overdispersion Model}

One of the first models for identifying somatic mutations in this clonal sequencing setting (without germline references) \citep{Spencer-Chapman2021-cy} uses the observation that for truly somatic variants, only a fraction of sampled clonal lineages may carry it, making the \textit{pooled} alternative read count ($R$) across all samples \textit{overdispersed}. One way to capture this by estimating the overdispersion ($\hat{\rho}$) of a beta-binomial distribution accounting for the reads:

$$
R | N \sim BetaBinom\left(\hat{p}\frac{1 - \rho}{\rho}, (1 - \hat{p})\frac{1-\rho}{\rho}\right)
$$

, where $\hat{p}$ is the MLE estimate for the VAF. In the original work\citep{Spencer-Chapman2021-cy}, the proposal was simply to exclude any variant with $\hat{\rho} > 0.1$. This approach works quite well in simulations with fairly even coverage across samples (at the proposed variant in question). In \texttt{pGermlinePoly} we provide this ``pooled overdispersion'' estimator as a secondary annotation under request in case the user would like to incorporate this in the future.

\bibliographystyle{unsrt} % We choose the "plain" reference style
\bibliography{refs} % Entries are in the refs.bib file


\end{document}
